Tree establishment in models of forest dynamics: a key priority for further research

### 2022-12-14 manuscript/manuscriptRW.Rmd
updates in the manuscript

---
title: "(WT1) Tree establishment in models of forest dynamics: a key priority for further research  \n(WT2) Tree establishment processes in models of forest dynamics: capturing the drivers remains challenging"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001]. All these processes are difficult to
assess, and some of them
are scarcely understood and thus appear highly stochastic
[@bogdziewicz2021]. Identifying the appropriate level of complexity for
the formulation of the key factors that are leading to successful tree
establishment is challenging. Currently, tree
establishment processes in dynamic forest models are handled in a
multitude of
ways [@konig_tree_2022; @bugmann2022], from entirely ignoring it [as
done in classical forest growth models, e.g., Pretzsch et al.
-@pretzsch2002], across the use of a few simple environmental filters,
as done in most forest \"gap\" models [@shuman2011] and Dynamic Global
Vegetation Models [e.g., @hickler2012; @smith2001], to complex
approaches that incorporate local feedback from the canopy, multiple
ecological processes and often also small time steps [e.g., @seidl2012;
@wehrli2006], or field-based statistical parameterizations, which
however are not easy to extrapolate in
space and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest
dynamics to represent tree establishment, model behavior is often facing
problems that can have strong impacts on simulated forest dynamics, such
as very high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to arrive at correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
\"classical\" forest gap models [@gutiérrez2016], at least as long as
the simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to
unrealistic successional drift in the model, which must be corrected by
factors that are hard to parameterize [e.g., @lischke2006]. A related
issue is the excessive reduction of species richness due to positive
feedback effects such that eventually just single-species ~~forest~~
stands remain. This is sometimes corrected by the incorporation of a low
level of seed influx of all species at all times [@lischke2006].
However, simulated species composition is usually exceedingly sensitive
to that level, whereas the parameters of such functions are poorly
constrained by field data. Lastly, there is often a problem with
insufficient observational constraint on parameter values for models
that start from very small tree sizes (e.g., 10 cm of height or even
seeds) and track tree development in a process-oriented manner by
considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

@@

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold (\"ingrowth\")
and thus as successful tree establishment. We included models that are
based on a range of \"philosophies\", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to \"process-based\" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not included in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.

@@

Fifteen models of forest dynamics (including two model featuring
variants) were used to assess tree establishment. They simulate forest
dynamics at the stand, landscape and global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely \"process-based\" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SYBILA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and
LandClim)[,]{.underline} whereas some [a]{.underline}re based on plant
ecophysiological processes (e.g., FORMIND and iLand). The different
approaches underlying the models have strongly influenced the
formulation of tree establishment.

The establishment modules can be categorized as \'regeneration\' or
\'recruitment\'
models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination
seedling [g]{.underline}rowth, which ultimately lead to the simulated
number of trees. Recruitment models, in contrast, introduce a certain
number of new trees with certain characteristics such as biomass or
diameter, without explicitly considering earlier development processes.
We can further distinguish models that feature a feedback in the
establishment, i.e., where the level and species composition
of establishment is influenced by the existing adult tree community via
the
production of seeds, seedlings or saplings, from models that do not
contain such feedback.

@@

Two models, ForClim and ForCEEPs, were used here with two alternative
variants.
ForClim variant 1 [@bugmann1996] is based on a recruitment module that
adheres closely to the concept introduced by @botkin1972, whereas
ForClim variant 11 is adopting a rather different approach that
considers space filling along with different \'points of entry\' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback via seed trees for the
quantity and quality of simulated ingrowth[,]{.underline} as one variant
includes this feedback whereas the other does not.

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a diameter threshold of 7 or 10
cm[,]{.underline} respectively (synonym: ingrowth). To this end, each
modeling group was provided with a detailed protocol (cf. Supplementary
Material 1, henceforth \"SM\") with instructions how to perform the
simulations, providing input variables on climate and soil conditions,
and the expected output variables. Neither were further site information
(except for the data specified below) nor any data on tree ingrowth
provided prior to the simulation. That is, the models were run in
\"blind flight\" mode.

@@

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1-ha plot in the equilibrium over time, or (2)
sampling several 1-ha plots from the simulated forest at one specific
point in time (in the equilibrium). This resulted in 880.000
observations per model that simulated the 11 species included in the
protocol (200 sites, 200 samples per site, 11 species and two diameter
thresholds). For the models that simulated additional species, their
ingrowth rates were aggregated as \"others\".

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 100 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management PLEASE **4C** VERIFY THIS IS
WHAT YOU DID. LPJ-GUESS had 2% of the sample outputs as grasslands that
were considered in the data as zero stand basal area and zero ingrowth
for all the tree species simulated. LPJ-GUESS could not simulate one of
the sites HERE **LPJ-GUESS** PROVIDES [ONE SENTENCE]{.underline} WHY
THEY DID NOT PROVIDE ALL THE SITES. Landis-II also missed X% of the
samples HERE **LANDIS-II** PROVIDES [ONE SENTENCE]{.underline} WHY THEY
DID NOT PROVIDE ALL THE SAMPLES.

@@

The fact that some models featured lower mortality than expected by
Reineke\'s self-thinning rule may point to facilitation, or simply a
higher mortality rate before the trees had reached 7
cm diameter. Yet, the case of models such as xComp, PICUS and TreeMig,
whose mortality was well above the self-thinning line, indicates that
they feature higher ingrowth mortality in mixed-species forests compared
to even-aged, single-species forests. Some models yielded a mortality
rate of ingrowth that is broadly compatible with the self-thinning rule
(i.e., 4C and LandClim).

@@

The observed data that we used [@käber2022] are also exceptional,
particularly with respect to the number of events captured, which is
essential for a highly \'noisy\' process such as tree establishment. The
use of this dataset for model benchmarking provided novel insights on
the capability of state-of-the art models to simulate ingrowth levels,
species composition, and mortality in an early stage of tree life, i.e.,
between a tree diameter of 7 and 10 cm. Overall, by adopting this
approach a much broader understanding resulted than if we had used these
data for model calibration: it is primarily from the failure of the
models that we can gain ecological insights [e.g., @trugman2022].

@@

From a modeling point of view, excess ingrowth inevitable requires
excess mortality rates
in a later stage, either -- as observed for some models -- between the 7
and 10 cm diameter thresholds, before the 7 cm are reached, or soon
after the 10 cm threshold has been crossed. In any case, correcting at
early stages for the expected forest densities at later stages is
equivalent to compensating for a first error (excess ingrowth) by a
second error (excess mortality). Thus, many models appear to get the
\'right answer\', i.e., correct stand structure and composition, for the
wrong reasons. It is highly likely that biased projections will result,
because the two errors are unlikely to be perfectly linked and thus will
not always compensate each other. Hence, this structural problem of most
models investigated here is problematic particularly if the models are
to be used under novel conditions such as under climate change [e.g.,
@huber2021] or in a decision support context [e.g., @thrippleton2021].

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001]. All these processes are difficult to
assess, and some of them are scarcely understood and thus appear highly
stochastic [@bogdziewicz2021]. Identifying the appropriate level of
complexity for the formulation of the key factors that are leading to
successful tree establishment is challenging. Currently, tree
establishment processes in dynamic forest models are handled in a
multitude of ways [@konig_tree_2022; @bugmann2022], from entirely
ignoring it [as done in classical forest growth models, e.g., Pretzsch
et al. -@pretzsch2002], across the use of a few simple environmental
filters, as done in most forest "gap" models [@shuman2011] and Dynamic
Global Vegetation Models [e.g., @hickler2012; @smith2001], to complex
approaches that incorporate local feedback from the canopy, multiple
ecological processes and often also small time steps [e.g., @seidl2012;
@wehrli2006], or field-based statistical parameterizations, which
however are not easy to extrapolate in space and time [e.g.,
@ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, and (2) assess future tree establishment
e.g., in the context of global change scenarios. Given the current
strategies that are used in models of forest dynamics to represent tree
establishment, model behavior is often facing problems that can have
strong impacts on simulated forest dynamics, such as very high levels of
tree establishment, which then necessitate excess mortality at early
stages of tree life to arrive at correct stand structure and
composition. Also, inadequately high species diversity in tree
establishment may be simulated. This is characteristic of "classical"
forest gap models [@gutiérrez2016], at least as long as the simulations
comprise a multitude of species. Other models use calibration against
local canopy-level data to define appropriate ingrowth levels, which is
likely to strongly hamper the general applicability of these models
e.g., under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species richness due to positive feedback effects such that eventually
just single-species ~~forest~~ stands remain. This is sometimes
corrected by the incorporation of a low level of seed influx of all
species at all times [@lischke2006]. However, simulated species
composition is usually exceedingly sensitive to that level, whereas the
parameters of such functions are poorly constrained by field data.
Lastly, there is often a problem with insufficient observational
constraint on parameter values for models that start from very small
tree sizes (e.g., 10 cm of height or even seeds) and track tree
development in a process-oriented manner by considering a multitude of
ecological influences, rather than emphasizing tree establishment into a
larger size class [e.g., @wehrli2006].

@@

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth")
and thus as successful tree establishment. We included models that are
based on a range of "philosophies", from models operating at the stand
to the global scale as well as the range of models from empirically
derived to "process-based" [e.g., @bugmann1996; @fabrika2012;
@mette2014; @lexer2001]. Because of the large variability in tree
establishment patterns in nature and the large number of factors driving
this process --- including some that are not included in most models
such as deer browsing --- we do not aim at a detailed statistical
evaluation of each model. Instead, we aim at evaluating the general
ingrowth patterns and magnitudes simulated by the models, and benchmark
the simulated regeneration niche of multiple species against empirical
data along a wide environmental gradients of temperature, moisture, and
light availability.

@@

Fifteen models of forest dynamics (including two model featuring
variants) were used to assess tree establishment. They simulate forest
dynamics at the stand, landscape and global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SYBILA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and
LandClim)[,]{.underline} whereas some [a]{.underline}re based on plant
ecophysiological processes (e.g., FORMIND and iLand). The different
approaches underlying the models have strongly influenced the
formulation of tree establishment.

The establishment modules can be categorized as 'regeneration' or
'recruitment' models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination
seedling [g]{.underline}rowth, which ultimately lead to the simulated
number of trees. Recruitment models, in contrast, introduce a certain
number of new trees with certain characteristics such as biomass or
diameter, without explicitly considering earlier development processes.
We can further distinguish models that feature a feedback in the
establishment, i.e., where the level and species composition of
establishment is influenced by the existing adult tree community via the
production of seeds, seedlings or saplings, from models that do not
contain such feedback.

@@

Two models, ForClim and ForCEEPs, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback via seed trees for the
quantity and quality of simulated ingrowth[,]{.underline} as one variant
includes this feedback whereas the other does not.

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a diameter threshold of 7 or 10
cm[,]{.underline} respectively (synonym: ingrowth). To this end, each
modeling group was provided with a detailed protocol (cf. Supplementary
Material 1, henceforth "SM") with instructions how to perform the
simulations, providing input variables on climate and soil conditions,
and the expected output variables. Neither were further site information
(except for the data specified below) nor any data on tree ingrowth
provided prior to the simulation. That is, the models were run in "blind
flight" mode.

@@

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1-ha plot in the equilibrium over time, or (2)
sampling several 1-ha plots from the simulated forest at one specific
point in time (in the equilibrium). This resulted in 880.000
observations per model that simulated the 11 species included in the
protocol (200 sites, 200 samples per site, 11 species and two diameter
thresholds). For the models that simulated additional species, their
ingrowth rates were aggregated as "others".

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 100 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management PLEASE **4C** VERIFY THIS IS
WHAT YOU DID. LPJ-GUESS had 2% of the sample outputs as grasslands that
were considered in the data as zero stand basal area and zero ingrowth
for all the tree species simulated. LPJ-GUESS could not simulate one of
the sites HERE **LPJ-GUESS** PROVIDES [ONE SENTENCE]{.underline} WHY
THEY DID NOT PROVIDE ALL THE SITES. Landis-II missed 0.2% of the
observations HERE **LANDIS-II** PROVIDES [ONE SENTENCE]{.underline} WHY
THEY DID NOT PROVIDE ALL THE SAMPLES/observations.

@@

The fact that some models featured lower mortality than expected by
Reineke's self-thinning rule may point to facilitation, or simply a
higher mortality rate before the trees had reached 7 cm diameter. Yet,
the case of models such as xComp, PICUS and TreeMig, whose mortality was
well above the self-thinning line, indicates that they feature higher
ingrowth mortality in mixed-species forests compared to even-aged,
single-species forests. Some models yielded a mortality rate of ingrowth
that is broadly compatible with the self-thinning rule (i.e., 4C and
LandClim).

@@

The observed data that we used [@käber2022] are also exceptional,
particularly with respect to the number of events captured, which is
essential for a highly 'noisy' process such as tree establishment. The
use of this dataset for model benchmarking provided novel insights on
the capability of state-of-the art models to simulate ingrowth levels,
species composition, and mortality in an early stage of tree life, i.e.,
between a tree diameter of 7 and 10 cm. Overall, by adopting this
approach a much broader understanding resulted than if we had used these
data for model calibration: it is primarily from the failure of the
models that we can gain ecological insights [e.g., @trugman2022].

@@

From a modeling point of view, excess ingrowth inevitable requires
excess mortality rates in a later stage, either -- as observed for some
models -- between the 7 and 10 cm diameter thresholds, before the 7 cm
are reached, or soon after the 10 cm threshold has been crossed. In any
case, correcting at early stages for the expected forest densities at
later stages is equivalent to compensating for a first error (excess
ingrowth) by a second error (excess mortality). Thus, many models appear
to get the 'right answer', i.e., correct stand structure and
composition, for the wrong reasons. It is highly likely that biased
projections will result, because the two errors are unlikely to be
perfectly linked and thus will not always compensate each other. Hence,
this structural problem of most models investigated here is problematic
particularly if the models are to be used under novel conditions such as
under climate change [e.g., @huber2021] or in a decision support context
[e.g., @thrippleton2021].
